---
title: "lmmvar-vignette"
output: rmarkdown::html_vignette
bibliography: REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{lmmvar-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This document provides a short tutorial on how to use the `lmmvar` package
which computes confidence regions and intervals in a linear mixed model with one
random effect, or a variance component model with one variance component and 
error term. The methods are based on the score and information matrix of the 
restricted likelihood. Theoretical motivations and additional details are in the
paper "Fast and reliable confidence intervals for a variance component" by
Zhang, Ekvall, and Molstad.

If you encounter any errors or strange behavior, please report it as an issue at
https://github.com/yqzhang5972/lmmvar.

## Installation
You can install the package from github using devtools:
```{r}
# install.packages("devtools")
# devtools::install_github("yqzhang5972/lmmvar")
library(lmmvar)
```


## Model
Suppose a vector of responses $y \in \mathbb{R}^n$ and design matrices 
$X\in\mathbb{R}^{n\times p}$ and $Z \in \mathbb{R}^{n\times q}$ satisfy, for some
$\beta \in \mathbb{R}^p$,
$$
y = X\beta + ZU + E,
$$
where $U \sim N(0, \sigma_g^2I_q)$ and $E \sim N(0, \sigma_e^2I_n)$ are independent.
Equivalently, with $K = ZZ^\top$, $\sigma^2_p = \sigma^2_g + \sigma^2_e$
and $h^2 = \sigma^2_g / \sigma^2_p$,

$$
y \sim N\{X\beta, \sigma_p^2(h^2K+(1-h^2)I_n)\}.
$$

The parameter $h^2$ have different names depending on the application, including
for example heritability or the proportion of spatial variability. Another common
parameter of interest is $\tau = \sigma^2_g / \sigma^2_e$, which is related to $h^2$
as $h^2 = \tau / (1 + \tau)$.

## Main functions
There are four main functions in the `lvmmr` package:

  - `varRatioTest1d`: Restricted score test-statistic for a proportion of variation $h^2$, or heritability.
  - `varRatioTest2d`: Joint restricted score test for a proportion of variation, or heritability, and total variation $(h^2, \sigma_p^2)$.
  - `confInv`:  Confidence interval for a proportion of variation $h^2$, or heritability.
  - `confReg`$:  Joint confidence region for proportion of variation, or heritability, and total variation $(h^2, \sigma_p^2)$.

We also provide functions based on restricted likelihood ratio test method (@crainiceanu2004likelihood) to construct confidence interval for the variance ratio $\tau = \sigma_g^2/\sigma_e^2$. The functions are

  - `simulate_RLRT`: Simulated distribution of the restricted likelihood ratio test statistic of variance ratio $\tau = \sigma_g^2/\sigma_e^2$.
  - `RLRTCI`: Confidence interval for variance ratio $\tau = \sigma_g^2/\sigma_e^2$ based on inverting the restricted likelihood ratio test using
  sample quantiles from parametric boostrapo samples under the null hypothesis.


## Example
### Set up
Let's look at an simulation example. We set the true $h^2$ to be 0.5 and $\sigma_p^2$ to be 1, and $(n, p) = (500, 10)$.
Since the restricted likelihood does not depend on the true $\beta$, we set it
equal to zero for simplicity.
```{r}
# -----------------------------------   
# Simulate data
# ----------------------------------- 
set.seed(0)
n <- 500 
p <- 10
h2 <- 0.5
s2p <- 1
K <- 0.5^abs(outer(1:n, 1:n, "-")) # AR(1)
X <- matrix(rnorm(n * p), nrow = n)
Ksqrt <- t(chol(K)) # This ensures tcrossprod(Ksqrt) = K, so Z = Ksqrt

# This is our generated response vector
y <- Ksqrt %*% rnorm(n, sd = sqrt(h2)) + rnorm(n, sd = sqrt(1 - h2))
```

Now let's see how to use \texttt{confInv} to construct a confidence interval for $h^2$. The arguments to the function are 

  -  `y`: A vector of length n of observed responses with diagonal covariance matrix.
  -  `X`: An n-by-p matrix of predictors, n > p.
  -  `lambda`: A vector of length n with (non-negative) variances, which are the eigenvalues of the variance component covariance matrix.
  -  `tolerance`: A positive scalar with the tolerance used in bisection search, default is 1e-4.
  -  `confLevel`: A number in (0, 1) with the level of the confidence interval, default is 0.95.
  -  `maxiter`: A positive integer. Stop and warning if number of iterations in search exceeds this value, default is 50.
  -  `type`: A string that is either "two-sided", "lower_bd" (lower bound only) or "upper_bd" (upper bound only), default is "two-sided".

The function \texttt{confInv} requires the argument `y` to have diagonal covariance matrix
$\sigma^2_p\{h^2\Lambda + (1 - h^2)I_n\}$. This is not true for the `y` we generated,
but we can transform it to satisfy the requirement: if $K = V\Lambda V^\top$ by
eigendecomposition, then $$V^\top y \sim N\{V^\top X\beta, \sigma_p^2(h^2\Lambda+(1-h^2)I_n)\}.$$

```{r}
# ------------------------------------------------------------------------------
# Transform X and y to make the covariance matrix diagonal
# ------------------------------------------------------------------------------
eigen_decomp = eigen(K, symmetric = T)
V = eigen_decomp$vectors               
lambda = eigen_decomp$values

X = crossprod(V, X)
y = crossprod(V, y)
```


### Score-based confidence interval
With the transformed data, we can calculate a confidence interval for $h^2$:
```{r}
confInv(y, X, lambda)
```

### Score test-statistic

If rather than to compute a confidence interval, we wanted to test a particular
null hypothesis, we can evaluate the test-statistic for $h^2$, or the joint test-
statistic for $(h^2, \sigma^2_p)$ at the null hypotheses:
```{r message=FALSE}
# Score test-statistic of h2 at true value
varRatioTest1d(h2, y, X, lambda)   
# Score test-statistic of (h2,sigma^2_p) at true value
varRatioTest2d(h2, s2p, y, X, lambda) 
```

### Joint confidence region
There is also a function to find test statistics for $(h^2, \sigma_p^2)$ evaluated
on a grid. The default grid range is from 0 to 1 for both $h^2$ and $\sigma_p^2$.
The output of this function is a matrix containing the test-statistic evaluated
at each grid point, which can be used to find a confidence region as below.
```{r message=FALSE, fig.width=6, fig.height=4}
range_h <- c(0.3, 0.95)
range_p <- c(0.8, 1.5)
n_grid <- 50
cr <- confReg(y, X, lambda, range_h = range_h, range_p = range_p, grid = n_grid)  
dim(cr)

# Plot the contours of the test-statistic
h2_vals <- seq(from = range_h[1], to = range_h[2], length.out = n_grid)
s2p_vals <- seq(from = range_p[1], to = range_p[2], length.out = n_grid)
contour(x = s2p_vals, y = h2_vals, z = cr,
        levels = qchisq(c(0.9, 0.95, 0.99), df = 2), # Conf. levels contours
        labels = c(0.9, 0.95, 0.99),
        ylab = latex2exp::TeX("$h^2$"),
        xlab = latex2exp::TeX("$\\sigma^2_p$"))

```

### Confidence interval using the restricted likelihood ratio
This function gives confidence interval for $\tau = \sigma^2_g / \sigma^2_e$ based
on RLRT method (@crainiceanu2004likelihood). First we need to find simulated
distribution of RLRT statistic using simulate_RLRT. The input of simulate_RLRT is

  -  `X`: An n-by-p matrix of predictors, n > p.
  -  `Ksqrt`: A matrix which is the square root of the variance component
      covariance matrix.
  -  `tau0`: A positive scalar indicates the value at which we want to make the
      hypothesis test.
  
The output of simulate_RLRT is 

  -  A vector with the length of number of simulation.

To create a confidence interval, we need to simulate from the distribution
of the test-statistic, for every potential value of the parameter of interest.
We have set the `nsim` and `length.out` arguments somewhat low to decrease
computing time; in practice one can incraese these to increase precision.
```{r}
# The potential values of tau
tau_seq <- seq(from = 0, to = 10, length.out = 100)

# Simulate the distribution of the test-statistics for every value in tau_seq
SimDists <- list()
for (j in 1:length(tau_seq)) {
  SimDists[[j]] <- simulate_RLRT(X, Ksqrt, tau0 = tau_seq[j], nsim = 500)
}
tau_int <- RLRTCI(SimDists = SimDists, ciseq = tau_seq, X = X, y = y,
                  lambda = lambda)
# CI for tau
tau_int
# Corresponding interval for h2
tau_int / (1 + tau_int)
```

