---
title: "lmmvar-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lmmvar-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this document, we provide a short tutorial on how to use the $\texttt{lmmvar}$ (Fast and reliable confidence intervals for a variance
component or proportion) software in R. 
If you encounter any errors or strange behavior, please report it as an issue at https://github.com/yqzhang5972/lmmvar.

## model set-up
Suppose $y = X\beta + Zu + e$,
where $u_{m\times1} \sim N(0, \sigma_g^2/m)$, $e_{n\times1} \sim N(0, \sigma_e^2I_n)$, so that
$$y \sim N(X\beta, \sigma_g^2K+\sigma_e^2I_n)$$
with $K = ZZ^T/m$. 

Now we would like to estimate test statistic and confidence interval for $h^2 = \sigma_g^2 / (\sigma_g^2+\sigma_e^2)$ (heritability) and $\sigma_p^2 = \sigma_g^2+\sigma_e^2$.

## Installation
```{r}
# install.packages("devtools")
# library(devtools)
# devtools::install_github("yqzhang5972/lmmvar")
library(lmmvar)
```

## Example
```{r}
# simulate y = X\beta + Zu + e
simData <- function(sigma2_g, sigma2_e = 1, X, Z) {  
  n = dim(X)[1]
  p = dim(X)[2]
  m = dim(Z)[2] 
  beta = rep(0, p)
  e = rnorm(n, mean = 0, sd = sqrt(sigma2_e))
  u = rnorm(m, mean = 0, sd = sqrt(sigma2_g/m))
  y = X %*% beta + Z %*% u + e
  return(y)
}
# simulate Z
simZ <- function(m,n) {
  S <- matrix(0, nrow = n, ncol = n)
  for (kk in 1:n) {
    for (jj in 1:n) {
      S[jj,kk] <- 0.9 ^ abs(jj - kk)
    }
  }
  eo <- eigen(S)
  SSqrt <- eo$vec %*% diag(eo$val^0.5) %*% t(eo$vec)
  Z <- SSqrt %*% matrix(rnorm(n*m), nrow = n)
  return(Z)
}


N = 10000
n=100
p=10
m=1000
h2 = 0.5
s2p = 1 

set.seed(0)
X = matrix(rnorm(n*p, 0, 1), nrow = n)   # simulate a n by p design matrix
Z <- simZ(m, n) # simulate z
y = simData(sigma2_g = h2*s2p, sigma2_e = (1-h2)*s2p, X, Z)  # simulate y = X\beta + Zu + e
K = Z %*% t(Z) / m  

eigen_decomp = eigen(K, symmetric = T) # eigen-decomposition of K
V = eigen_decomp$vectors       # finding K's eigenvectors
lambda = eigen_decomp$values   # finding K's eigenvalues
Xnew = crossprod(V, X)         # transform X with V'X
ynew = crossprod(V, y)         # transform y with V'y
```

## Compute the 1d and 2d test statistics, 1d confidence interval and 2d test statistics matrix
```{r message=FALSE}
# score test statistic of h2 at true value
varRatioTest1d(h2, ynew, Xnew, lambda)   
# score test statistic of (h2,sigma^2_p) at true value
varRatioTest2d(h2, s2p, ynew, Xnew, lambda) 
# 95% confidence interval of h2 treating sigma^2_p as nuisance parameter
confInv(ynew, Xnew, lambda) 

# matrix of test statistics at a grid of h2 from 0 to 1 and sigma^2_p from 0.0001 to 3.
cr <- confReg(range_p = c(1e-4,3), ynew, Xnew, lambda) 
library(ggplot2)
library(dplyr)
library(latex2exp)
data.frame(p = seq(1e-4,3, length.out = 200)[30:150], teststatistic = cr[100,30:150]) %>%
  ggplot(aes(x = p, y = teststatistic))  + theme_bw() + #geom_point(size=0.8) +
  geom_line() + xlab(TeX("$sigma^2_p$")) + ylab("Test-statistic") +
  ggtitle(TeX("$h^2=0.5$")) + theme(plot.title = element_text(hjust = 0.5)) +
  geom_hline(yintercept=qchisq(0.95, 1), color = "red")+
  theme(axis.title=element_text(size=16), 
        plot.title = element_text(size = 18), axis.text = element_text(size=12))
```

